Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-28 papers/wdmerger_I/paper.tex
Use the emulateapj class for this preprint.

\documentclass[12pt,preprint]{aastex}
% for \sout
\usepackage{ulem}
% makes sure \em{} is italic rather than underlined (corrects ulem from line above)
\normalem

@@

@@

\shorttitle{DWD Mergers. I. Methodology}
\shortauthors{Max}

@@

\citet{marcello:2012} use the mass continuity equation to rewrite the
standard source term from Equation
\ref{eq:cell_center_gravity_source}. When this is done, the resulting
evolution equation for the total energy is:
\begin{equation}
  \frac{\partial (\rho E_{\text{tot}})}{\partial t} = -\bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) + \frac{1}{2}\phi\frac{\partial \rho}{\partial t} - \frac{1}{2}\rho\frac{\partial \phi}{\partial t}.
\end{equation}
In practice, we would still actually evolve the total gas energy
$E_G$, and so when discretized to second order in time from time level
$n$ to time level $n+1$ the gas energy equation would look like
\begin{align}
  (\rho E_G)^{n+1} &= (\rho E_G)^{n} + \left(\frac{1}{2} (\rho \phi)^{n+1} - \frac{1}{2} (\rho \phi)^{n}\right) - \Delta t\, \bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) \notag \\
   &+ \frac{\Delta t}{2}\left(\phi \frac{\partial \rho}{\partial t}\right)^{n+1/2} - \frac{\Delta t}{2} \left(\rho \frac{\partial \phi}{\partial t}\right)^{n+1/2}.
\end{align}
It is not immediately obvious that this is explicitly conservative for
the total energy, due to the last two terms on the right-hand side,
but written in discrete form the potential is linked to the density in
such a way that there is indeed a perfect cancellation of terms
between adjacent cells (see Equation 88 in
\citeauthor{marcello:2012}). Doing it this way requires a prescription
for calculating $\phi^{n+1/2}$, the potential at the time halfway
between the new and old state, as well as $\dot{\phi}^{n+1/2}$, the
time derivative of the potential at that time. To second order in
time, these can be calculated as $(\phi^{n+1} + \phi^{n})/2$ and
$(\phi^{n+1}-\phi^{n})/\Delta t$ respectively. So, in practice what we
do is to first solve the Poisson equation using $\rho^n$ at the
beginning of a time step; then, evolve the pure hydrodynamics
equations without the gravitational source to obtain $\rho^{n+1}$;
then, solve the Poisson equation for $\phi^{n+1}$; and then to finally
calculate the gravitational source terms and add them in. An
alternative approach is to use the predictor-corrector scheme that
CASTRO features, where the time level $n$ source term is included in
the hydrodynamics update and then left out of the final update. In
either case, the time level $n$ calculation of $\mathbf{g}$ is
included the PPM reconstruction of the edge states.

@@

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = G\, \sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{direct_sum}
\end{equation} \MarginPar{should there really be 2 $G$'s?}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,
m, n)$ \MarginPar{indices should be in normal math style, not roman}
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation
\ref{spherical_harmonic_expansion} using the addition theorem for
spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty} \left[Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} + \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right].\label{multipole_potential}
\end{equation}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the
outer summation in Equation \ref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as \MarginPar{We usually use $\| \ldots \|_2$ for a norm}
\begin{equation}
  L^2(f) = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{L^2(\phi - \phi_{\text{sphere}})}{L^2(\phi_{\text{sphere}})}.
\end{equation}

@@

The only way to fully eliminate this effect is to test a problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \phi_{\text{cube}}(x,y,z) &= G\rho\sum_{i,j,k=0}^1\left(x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right) + y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) - \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right)
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error norm for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and indeed we see
near-perfect second-order scaling.

@@

@@

Following \cite{robertson:2010}, we set up a Kelvin-Helmholtz test in
the following way. The problem domain runs from 0 to 1 in both the $x$
and $y$ directions. Though this is a two-dimensional test, we run
CASTRO in 3D and replicate the problem along 8 cells in the $z$
direction; the physical size of these zones is determined so that $dz
= dx = dy$, a requirement at present in CASTRO. The problem involves a
fluid slab of density $\rho_1 = 2.0$ traveling rightward in the
$x$-direction at velocity $v_1 = 0.5$, sandwiched by a fluid of
density $\rho_2 = 1.0$ traveling leftward at velocity $v_2 =
-0.5$. The density gradient is in the $y$ direction, so this creates a
velocity shear along the interface between the fluids. The density and
velocity distribution on the computational domain are given by:
\begin{align}
  \rho &= \rho_1 + R(y)\left[\rho_2 - \rho_1\right] \\
  v_x  &= v_1 + R(y)\left[v_2 - v_1\right] \\
  v_y  &= v_{\text{bulk}} + v^\prime
\end{align}
Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem will be established for two
sets of initial conditions (ICs), which we follow
\citeauthor{robertson:2010} in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \\
  R_B &= \Big\{\left[1 + \text{exp}(-2(y-0.25)/\Delta_y)\right]\left[1 + \text{exp}(2(y-0.75)/\Delta_y)\right]\Big\}^{-1} \\
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{\text{exp}\left[-\frac{(y-0.25)^2}{2\sigma^2}\right] + \text{exp}\left[-\frac{(y-0.75)^2}{2\sigma^2}\right]\right\} \\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right).
\end{align}
Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta_y$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.01 seconds, and the problem is run until $t = 2$.

@@

\begin{figure}
  \centering
  \includegraphics[scale=2.0]{freefall/plot_freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the total elapsed time
    necessarily. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO. The agreement between
    theory and simulation is excellent, even for a simulation with
    modest resolution.}
  \label{Fig:Free Fall}
\end{figure}

>>>>

%\documentclass[12pt,preprint]{aastex}
\documentclass{emulateapj}
% for \sout
\usepackage{ulem}
% makes sure \em{} is italic rather than underlined (corrects ulem from line above)
\normalem

@@

% allows comment sections
\usepackage{verbatim}

@@

\shorttitle{DWD Mergers. I. Methodology}
\shortauthors{Katz et al. (2015)}

@@

\citet{marcello:2012} use the mass continuity equation to rewrite the
standard source term from Equation
\ref{eq:cell_center_gravity_source}. When this is done, the resulting
evolution equation for the total energy is:
\begin{equation}
  \frac{\partial (\rho E_{\text{tot}})}{\partial t} = -\bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) + \frac{1}{2}\phi\frac{\partial \rho}{\partial t} - \frac{1}{2}\rho\frac{\partial \phi}{\partial t}.
\end{equation}
In practice, we would still actually evolve the total gas energy
$E_G$, and so when discretized to second order in time from time level
$n$ to time level $n+1$ the gas energy equation would look like
\begin{align}
  (\rho E_G)^{n+1} &= (\rho E_G)^{n} + \left(\frac{1}{2} (\rho \phi)^{n+1} - \frac{1}{2} (\rho \phi)^{n}\right) \notag \\
   &- \Delta t\, \bm{\nabla}\cdot\left(\rho \mathbf{u} E_{\text{tot}} + p\mathbf{u} - \frac{1}{2}\rho\mathbf{u}\phi\right) \notag \\
   &+ \frac{\Delta t}{2}\left(\phi \frac{\partial \rho}{\partial t}\right)^{n+1/2} - \frac{\Delta t}{2} \left(\rho \frac{\partial \phi}{\partial t}\right)^{n+1/2}.
\end{align}
It is not immediately obvious that this is explicitly conservative for
the total energy, due to the last two terms on the right-hand side,
but written in discrete form the potential is linked to the density in
such a way that there is indeed a perfect cancellation of terms
between adjacent cells (see Equation 88 in
\citeauthor{marcello:2012}). Doing it this way requires a prescription
for calculating $\phi^{n+1/2}$, the potential at the time halfway
between the new and old state, as well as $\dot{\phi}^{n+1/2}$, the
time derivative of the potential at that time. To second order in
time, these can be calculated as $(\phi^{n+1} + \phi^{n})/2$ and
$(\phi^{n+1}-\phi^{n})/\Delta t$ respectively. So, in practice what we
do is to first solve the Poisson equation using $\rho^n$ at the
beginning of a time step; then, evolve the pure hydrodynamics
equations without the gravitational source to obtain $\rho^{n+1}$;
then, solve the Poisson equation for $\phi^{n+1}$; and then to finally
calculate the gravitational source terms and add them in. An
alternative approach is to use the predictor-corrector scheme that
CASTRO features, where the time level $n$ source term is included in
the hydrodynamics update and then left out of the final update. In
either case, the time level $n$ calculation of $\mathbf{g}$ is
included the PPM reconstruction of the edge states.

@@

Analytic solutions to the Poisson equation customarily assume that the
potential vanishes at large distances from the region of non-zero
density. However, on a finite computational domain it is usually not
possible to have the edges of the domain be far enough away that the
potential can be taken to be zero there. Solving the Poisson equation
therefore requires knowledge of the values of the potential on the
edges of the computational domain. In principle, they can be computed
by doing a direct sum over the mass distribution inside the domain,
where the mass in each zone is treated as a point source:
\begin{equation}
  \Phi_{{lmn}} = \sum_{{i, j, k}} \frac{G \rho_{{ijk}}}{|\mathbf{x}_{{lmn}} - \mathbf{x}_{{ijk}}|}\, \Delta V_{{ijk}}.\label{direct_sum}
\end{equation}
Here $(i, j, k)$ are the indices of cells inside the domain, and $(l,m, n)$ 
are the indices of boundary locations. $\Delta V$ is the volume of the
zone. We have implemented this as an option\footnote{It is controlled
  with the \texttt{gravity.direct\_sum\_bcs} input parameter.} in
CASTRO. If there are $N$ zones per spatial dimension, then there are
$6 N^2$ boundary zones, and each boundary zone requires a sum over
$N^3$ zones, so the direct computation of the boundary conditions
scales as $\mathcal{O}(N^5)$.  This method is expensive enough that it is not used
for hydrodynamics simulations (though it is useful for comparison to
approximate solutions).

@@

The most natural extension of the monopole prescription is to include
higher-order multipole moments. If the entire mass distribution is
enclosed, then the potential can be expanded in a series of spherical
harmonics $Y_{lm}$:
\begin{equation}
  \Phi(\mathbf{x}) = \sum_{l=0}^{\infty}\sum_{m=-l}^{l} \frac{4\pi}{2l + 1} q_{lm} \frac{Y_{lm}(\theta,\phi)}{r^{l+1}}, \label{spherical_harmonic_expansion}
\end{equation}
where $q_{lm}$ are the so-called multipole moments. The origin of the
coordinate system is taken to be the center of the computational
domain, and $r$ is the distance to the origin. The multipole moments
can be calculated by expanding the Green's function for the Poisson
equation as a series of spherical harmonics.
\begin{comment}
, which yields
\begin{equation}
  q_{lm} = \int Y^*_{lm}(\theta^\prime, \phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3x^\prime. \label{multipole_moments_original}
\end{equation}
\end{comment}
After some algebraic simplification of Equation
\ref{spherical_harmonic_expansion} using the addition theorem for
spherical harmonics,
\begin{comment}
\begin{align}
  &\frac{4\pi}{2l+1} \sum_{m=-l}^{l} Y^*_{lm}(\theta^\prime,\phi^\prime)\, Y_{lm}(\theta, \phi) = P_l(\text{cos}\, \theta) P_l(\text{cos}\, \theta^\prime) \notag \\
  &\ \ + 2 \sum_{m=1}^{l} \frac{(l-m)!}{(l+m)!} P_{l}^{m}(\text{cos}\, \theta)\, P_{l}^{m}(\text{cos}\, \theta^\prime)\, \left[\text{cos}(m\phi)\, \text{cos}(m\phi^\prime) + \text{sin}(m\phi)\, \text{sin}(m\phi^\prime)\right].
\end{align}
\end{comment}
the potential outside of the mass distribution can be written as:
\begin{align}
  \Phi(\mathbf{x}) &= \sum_{l=0}^{\infty} \left\{Q_l^{(0)} \frac{P_l(\text{cos}\, \theta)}{r^{l+1}} \right. \notag \\
    &+ \left. \sum_{m = 1}^{l}\left[ Q_{lm}^{(C)}\, \text{cos}(m\phi) + Q_{lm}^{(S)}\, \text{sin}(m\phi)\right] \frac{P_{l}^{m}(\text{cos}\, \theta)}{r^{l+1}} \right\}.\label{multipole_potential}
\end{align}
$P_l(x)$ are the Legendre polynomials and $P_{l}^{m}(x)$ are the associated Legendre polynomials. $Q_l^{(0)}$ and $Q_{lm}^{(C,S)}$ are variants of the multipole moments that involve integrals of $P_l$ and $P_l^m$, respectively, over the computational domain.
\begin{comment}
\begin{align}
  Q_l^{(0)}   &= \int P_l(\text{cos}\, \theta^\prime)\, {r^{\prime}}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(C)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{cos}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime \\
  Q_{lm}^{(S)} &= 2\frac{(l-m)!}{(l+m)!} \int P_{l}^{m}(\text{cos}\, \theta^\prime)\, \text{sin}(m\phi^\prime)\, {r^\prime}^l \rho(\mathbf{x}^\prime)\, d^3 x^\prime.
\end{align}
\end{comment}
This approach becomes computationally feasible when we cut off the
outer summation in Equation \ref{multipole_potential} at some finite
value of $l_{\text{max}}$. If it is of sufficiently high order, we
will accurately capture the distribution of mass on the grid. In
practice we first evaluate the discretized analog of the modified
multipole moments for $0 \leq l \leq l_{\text{max}}$ and $1 \leq m
\leq l$, an operation that scales as $N^3$. We then directly compute
the value of the potential on all of the $6N^2$ boundary zones. Since
the multipole moments only need to be calculated once per Poisson
solve, the full operation scales only as $N^3$. The amount of time
required to calculate the boundary conditions will be directly related
to the chosen value of $l_{\text{max}}$, so there is a trade-off
between computational expense and accuracy of the result.

@@

Since the results of a merger simulation depend strongly on gravity,
it is important to check whether proper numerical convergence is
achieved for the Poisson solver. To do so, we created a simple test
that inserts a sphere of radius $R$ and uniform mass density $\rho$
onto our grid, and used CASTRO to calculate the gravitational
potential $\phi$ of this setup. We ensure that $R$ is an integer
multiple of the grid spacing, and the center of the sphere is at the
origin. The problem domain for our simulations is $[-1.6, 1.6]^3$, and
we take $R = 0.5$ and $\rho = 10^3$. The zones with $r > R$ are filled
with an ambient material of very low density. We run this problem at
multiple resolutions corresponding to jumps by a factor of two. For
comparison, at each grid point we evaluate the analytical potential of
a uniform sphere, which can be easily determined using Gauss' law:
\begin{equation}
  \phi_{\text{sphere}}(r) = \frac{GM}{r} \times \begin{cases} (3R^2 - r^2)/(2 r^2) & r \leq R \\ 1 & r > R \end{cases},\label{eq:sphere-analytical}
\end{equation}
where $M = 4\pi R^3 / 3$ is the mass of the sphere. Defining the $L^2$
norm of a field $f$ as
\begin{equation}
  \| f \|_2 = \left(\sum_{i,j,k} dx\, dy\, dz\, f_{ijk}^2\right)^{1/2},
\end{equation}
we measure the numerical error by calculating the $L^2$ norm of the error and normalizing it by the $L^2$ norm of the analytical solution:
\begin{equation}
  \text{Error} = \frac{\|\phi - \phi_{\text{sphere}}\|_2}{\|\phi_{\text{sphere}}\|_2}.
\end{equation}

@@

The only way to fully eliminate this effect is to test a problem that
does not change with resolution. The obvious companion is a cube of
uniform density $\rho$, where now $R$ is half of the side length of
the cube. At each resolution we use the same $R$ as for the sphere,
which ensures that the cube always fills exactly the same fraction of
the domain and thus has the same mass, so the only improvement comes
from better sampling at higher resolution. The potential for this
object has been worked out analytically by \citet{waldvogel:1976} (see
also a similar result in \citealt{hummer:1996}). The potential is in
Equation 15 of that paper, though the last term is missing a factor of
$1/2$, which destroys the symmetry. Inserting this missing factor and
performing a simple coordinate transformation so that the center of
the cube is at the origin, the potential is
\begin{align}
  \phi_{\text{cube}}(x,y,z) &= G\rho\sum_{i,j,k=0}^1\left[x_i y_j\, \text{tanh}^{-1}\left(\frac{z_k}{r_{ijk}}\right)\right. \notag \\
  &+ \left. y_j z_k\, \text{tanh}^{-1}\left(\frac{x_i}{r_{ijk}}\right) + z_k x_i\, \text{tanh}^{-1}\left(\frac{y_j}{r_{ijk}}\right) \right.\notag \\
  &\left. - \frac{x_i^2}{2}\,\text{tan}^{-1}\left(\frac{y_j z_k}{x_i r_{ijk}}\right) - \frac{y_j^2}{2}\,\text{tan}^{-1}\left(\frac{z_k x_i}{y_j r_{ijk}}\right) \right. \notag \\
  &- \left. \frac{z_k^2}{2}\,\text{tan}^{-1}\left(\frac{x_i y_j}{z_k r_{ijk}}\right)\right]
\end{align}
where $x_0 = R + x$, $x_1 = R - x$, $y_0 = R + y$, $y_1 = R-y$, $z_0 =
R+z$, $z_1 = R-z$, and $r_{ijk} = \sqrt{x_i^2 + y_j^2 + z_k^2}$. In
\texttt{FORTRAN} and \texttt{C}, the inverse hyperbolic tangent is
\texttt{atanh} and the inverse tangent is \texttt{atan} (\textit{not}
\texttt{atan2}). This formula is valid both inside and outside the
cube. The normalized $L^2$ error norm for this problem is also shown
in Figure \ref{fig:gravity_convergence}, and indeed we see
near-perfect second-order scaling.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{freefall/plot_freefall}
  \caption{Time evolution of two initially stationary white dwarfs,
    mutually attracted to each other by the gravitational force. The
    horizontal axis gives the separation of the white dwarfs, scaled
    to the initial separation, and the vertical axis gives the elapsed
    time of the simulation, scaled to the total elapsed time
    necessarily. The solid curve shows the analytical result,
    calculated from Newtonian mechanics, and the circles show the
    samples from the time evolution with CASTRO. The agreement between
    theory and simulation is excellent, even for a simulation with
    modest resolution.}
  \label{Fig:Free Fall}
\end{figure}

@@

Following \cite{robertson:2010}, we set up a Kelvin-Helmholtz test in
the following way. The problem domain runs from 0 to 1 in both the $x$
and $y$ directions. Though this is a two-dimensional test, we run
CASTRO in 3D and replicate the problem along 8 cells in the $z$
direction; the physical size of these zones is determined so that $dz
= dx = dy$, a requirement at present in CASTRO. The problem involves a
fluid slab of density $\rho_1 = 2.0$ traveling rightward in the
$x$-direction at velocity $v_1 = 0.5$, sandwiched by a fluid of
density $\rho_2 = 1.0$ traveling leftward at velocity $v_2 =
-0.5$. The density gradient is in the $y$ direction, so this creates a
velocity shear along the interface between the fluids. The density and
velocity distribution on the computational domain are given by:
\begin{align}
  \rho &= \rho_1 + R(y)\left[\rho_2 - \rho_1\right] \\
  v_x  &= v_1 + R(y)\left[v_2 - v_1\right] \\
  v_y  &= v_{\text{bulk}} + v^\prime
\end{align}
Here $R(y)$ is a ramp function that describes the transition between
the two fluids, while $v_{\text{bulk}}$ is the bulk motion of the
fluid in the $y$ direction and $v^\prime$ is the velocity perturbation
that seeds the instability. The problem will be established for two
sets of initial conditions (ICs), which we follow
\citeauthor{robertson:2010} in calling ICs A and B. They differ in
their ramp function ($R_A$ and $R_B$ respectively), as well as the
initial perturbation ($v^\prime_A$ and $v^\prime_B$ respectively), and
the frequency of the perturbation ($n_A = 4$ and $n_B = 2$):
\begin{align}
  R_A &= \begin{cases} 0 & |y - 0.5| > 0.25 \\ 1 & |y - 0.5| < 0.25 \end{cases} \\
  R_B &= \Big\{\left[1 + e^{-2(y-0.25)/\Delta_y}\right]\left[1 + e^{2(y-0.75)/\Delta_y}\right]\Big\}^{-1} \\
  v^\prime_A &= w_0\, \text{sin}\left(n_A\, \pi\, x\right) \left\{e^{-(y-0.25)^2 / 2\sigma^2} + e^{-(y-0.75)^2/2\sigma^2}\right\} \\
  v^\prime_B &= w_0\, \text{sin}\left(n_B\, \pi\, x\right).
\end{align}
Here $w_0 = 0.1$ is the scale of the velocity perturbation, $\sigma =
0.05/\sqrt{2}$ controls the width of the Gaussian for IC A, and
$\Delta_y$ is the transition distance scale for the smooth ramp of IC
B. The pressure everywhere is set to $p = 2.5$, and we run this with a
gamma-law equation of state set to $\gamma = 5/3$. Plotfiles are
generated every 0.01 seconds, and the problem is run until $t = 2$.

@@
